---
title: "Day 4 - R basics and functional coding"
author: "Lucas A. Nell and Maggie L. Warren"
date: "2023-11-27"
output:
  ioslides_presentation:
    incremental: false
    widescreen: true
    smaller: false
    css: classes-styles.css
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r source-html, results="asis", echo = FALSE}
# Sourcing file that includes extra html for our class lectures:
cat(readLines("class-preamble.html"), sep = "\n")
```



## Practice making tidy data

* In your group discuss what is wrong with the format of your data
* Create a repository for your metadata project
* Clone the repository to your local git
* Delegate jobs to group members:
    * write readme describing the data
    * reshape the data into a tidy form and change to csv
    * read the data into R making sure it is in the right form
* Each role will push changes to a new branch of the main branch
* As a group pull the changes merging the branches to the main branch
    

