---
title: "Day 5 - Super functional coding"
author: "Lucas A. Nell and Maggie L. Warren"
date: "2023-11-29"
output:
  ioslides_presentation:
    incremental: false
    widescreen: true
    smaller: false
    css: classes-styles.css
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
suppressPackageStartupMessages({
    library(tidyverse)
})
set.seed(1)
theme_set(theme_minimal() +
    theme(axis.line = element_line(linewidth = 1), 
          panel.grid = element_blank(), 
          axis.title = element_text(size = 14),
          axis.text = element_text(size = 12),
          axis.ticks = element_line(color = "black")))
```
```{r source-html, results="asis", echo = FALSE}
# Sourcing file that includes extra html for our class lectures:
cat(readLines("class-preamble.html"), sep = "\n")
```



## From last time: Cleaning weird files

We have a folder full of output files like this:

```{r n_slides_ex1, echo = FALSE}
n_slides <- 4
file_names <- paste0("weird-files/", c("file1.out", "file2.out"))
cat(read_lines(file_names[1]), sep = "\n")
```


```{r weirddata_data, echo = FALSE, eval=FALSE}
# I used this to create the files
files <- c("## Data provided by X

Ozone,Solar.R,Wind,Temp,Month,Day
41,190,7.4,67,5,1
NA,NA,14.3,56,5,5
--- instrument error
28,NA,14.9,66,5,6
23,299,8.6,65,5,7
--- instrument error
NA,194,8.6,69,5,10

## Year observed: 1990",
"## Data provided by Y

Ozone,Solar.R,Wind,Temp,Month,Day
7,NA,6.9,74,5,11
--- instrument error
11,290,9.2,66,5,13
14,274,10.9,68,5,14
18,65,13.2,58,5,15
--- instrument error
--- instrument error
6,78,18.4,57,5,18
30,322,11.5,68,5,19
11,44,9.7,62,5,20

## Year observed: 1991
")
```


## Cleaning weird files (2/`r n_slides`)

### Problems:

1. Remove unnecessary lines from each file.
2. Create a single data frame from multiple cleaned files.

### Input information:

1. Vector of file names (`file_names`)

```{r weirddata_info, eval=FALSE}
file_names <- c("file1.out", "file2.out")
```


### Get the files themselves from...

- Canvas (Files > 04_functionalR)
- <https://github.com/lucasnell/pranga/tree/main/weird-files>



## Cleaning weird files (3/`r n_slides`)

Clean a single CSV file to a string:

```{r weirddata3}
clean_str <- function(file_name) {
  lines <- read_lines(file_name)
  lines <- lines[!str_detect(lines, "^\\#\\#|^--")]
  lines <- lines[lines != ""]
  cleaned_str <- paste(lines, collapse = "\n")
  return(cleaned_str)
}
```



## Cleaning weird files (4/`r n_slides`)


Clean multiple files then combine them into a single data frame:

```{r weirddata4}
clean_df <- function(file_names) {
  cleaned_strs <- map(file_names, clean_str)
  data_frames <- map(cleaned_strs, read_csv, col_types = cols())
  combined_df <- bind_rows(data_frames)
  return(combined_df)
}
clean_df(file_names)
```



# Before we proceed...

## Let's create a global `.gitignore` file!<footnote>https://stackoverflow.com/a/7335487</footnote> {.smaller .smallcode}

### Mac, Linux, or Windows git bash:

```bash
git config --global core.excludesFile ~/.gitignore
```

### Windows cmd:

```bash
git config --global core.excludesFile "%USERPROFILE%\.gitignore"
```

### Windows Powershell:
```bash
git config --global core.excludesFile "$Env:USERPROFILE\.gitignore"
```

### Confirm the location:

```bash
git config --global core.excludesFile
```


## Add file with things to ignore

### Mac, Linux, or (maybe) Windows git bash:

```bash
touch ~/.gitignore
open ~/.gitignore
```

Otherwise, just open a file in a text editor and save it as `~/.gitignore`

Paste in the following contents to file:

<https://shorturl.at/jlLT9>

or Canvas > Files > 05_super-functional > gitignore.txt

Then save!



# Now back to functional coding


## Defensive programming {.build}

> "Defensive programming is a technique to ensure that code fails with
> well-defined errors, i.e. where you know it shouldn’t work.
> The key here is to 'fail fast' and ensure that the code throws an error as 
> soon as something unexpected happens.
> This creates a little more work for the programmer, but makes debugging 
> code a lot easier at a later date." <footnote>*Reproducible Code*, p 14 found at <https://www.britishecologicalsociety.org/publications/better-science/></footnote>


- Protect others that want to use your code (including your PI) from themselves
- Protect yourself from future you


## How to program defensively

```{r defensive-code1, error = TRUE}
subtract <- function(x, y) {
  z <- x - y
  return(z)
}
subtract("x", 6)
```

<br>
**Not very useful for debugging**



## How can we make this error message more useful? {.build}

Think about the necessary properties of your arguments

```{r defensive-code2, error = TRUE}
subtract <- function(x, y) {
  if (!is.numeric(x) || !is.numeric(y)) {
    stop("both x and y must be numeric")
  }
  z <- x - y
  return(z)
}
subtract("x", 6)
```

## An example of insidious R helpfulness {.build}

Sometimes R is just too helpful for its own good

What if we want to subtract `6:1` from `1:6`, but accidentally typed 
`6:5` instead of `6:1`?

```{r defensive-code3}
subtract(1:6, 6:5)
```

<br>
**What extra check do we need?**

## That extra check

```{r defensive-code4, error = TRUE}
subtract <- function(x, y) {
  if (!is.numeric(x) || !is.numeric(y)) {
    stop("both x and y must be numeric")
  }
  if (length(x) != length(y)) {
    stop("x and y must have the same length")
  }
  z <- x - y
  return(z)
}
subtract(1:6, 6:5)
```




## Sourcing files {.smaller .smallcode}

- `source("file.R")` runs all the code inside `file.R`
- It's pretty simple, but it can help a lot with coding organization
and staying DRY

<br>
Let's say we have a file `plot.R`:

```{r source-example1, eval = FALSE}
library(tidyverse)
my_df <- ChickWeight |> 
  as_tibble() |> 
  mutate(log_weight = log(weight))
my_df |> 
  ggplot(aes(Time, log_weight)) +
  ...
```

<br>
and another file `analyze.R`:

```{r source-example2, eval = FALSE}
library(tidyverse)
my_df <- ChickWeight |> 
  as_tibble() |> 
  mutate(log_weight = log(weight))
mod <- lm(log_weight ~ Time * Diet, my_df)
```


## Sourcing helps us avoid repeating code {.smaller .smallcode}

We can instead have three files:


<br>
`read-data.R`:

```{r source-example3, eval = FALSE}
library(tidyverse)
my_df <- ChickWeight |> 
  as_tibble() |> 
  mutate(log_weight = log(weight))
```

<br>
`plot.R`:

```{r source-example4, eval = FALSE}
source("read-data.R")
my_df |> 
  ggplot(aes(Time, weight)) +
  ...
```

<br>
`analyze.R`:

```{r source-example5, eval = FALSE}
source("read-data.R")
mod <- lm(log_weight ~ Time * Diet, my_df)
```

## Example of using `source()`

`scripts` folder inside
<https://github.com/lucasnell/gameofclones-data>


## How could you use this in your projects?



## Building R packages

First run this in R to get some useful packages:

```{r pkg-packages, eval = FALSE}
install.packages(c("devtools", "roxygen2", "testthat", "knitr"))
```

### Next, let's make a test package from scratch:

In R, run:

```{r create-pkg, eval = FALSE}
usethis::create_package("<path>/<pkg-name>")
```
where `<pkg-name>` should only contain letters, numbers, and periods

It should open up in RStudio



## Structure of our package


- `.Rbuildignore`: specifies files to avoid incorporating into the package.
  Use `usethis::use_build_ignore()` to add more files to it.
- `DESCRIPTION`: metadata about your package.
- `NAMESPACE`: contains exports from your package and imports from other 
  packages to yours. You shouldn't typically have to edit this.
- `R`: contains all the R code you're using for your package, and we'll use
  this to create our documentation, too. This is the main folder you'll work in.


## Hotkeys from RStudio


| Description               | Windows/Linux     | Mac                |
|:--------------------------|:------------------|:-------------------|
| Build and Reload	        | Ctrl+Shift+B      | Cmd+Shift+B        |
| Document Package	        | Ctrl+Shift+D      | Cmd+Shift+D        |
| Insert Roxygen Skeleton	  | Ctrl+Alt+Shift+R  | Cmd+Option+Shift+R |
| Test Package	            | Ctrl+Shift+T      | Cmd+Shift+T        |
| Check Package	            | Ctrl+Shift+E      | Cmd+Shift+E        |

## `DESCRIPTION`

Yours should look like this:

```
Package: testPkg
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person("First", "Last", , "first.last@example.com", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: `use_mit_license()`, `use_gpl3_license()` or friends to pick a
    license
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.3
```

## `DESCRIPTION` - What does your package do?

### `Title`

Short (<65 chars) and title case, does not end in period

### `Description`

Short paragraph with greater description, each line < 80 characters wide 
with lines after the first indented by 4 spaces.

From `ggplot2`:

```
Title: Create Elegant Data Visualisations Using the Grammar of Graphics
Description: A system for 'declaratively' creating graphics,
    based on "The Grammar of Graphics". You provide the data, tell 'ggplot2'
    how to map variables to aesthetics, what graphical primitives to use,
    and it takes care of the details.
```


## `DESCRIPTION` - `Version` field

Typically I make mine as `<major>.<minor>.<patch>`

1. The major version is only incremented when
   there are huge changes that likely affect most users.
2. The minor version is incremented for bug fixes and
   new features that are backward compatible.
3. The patch version fixes small bugs that have little change overall to the user.


- I add another integer and a `"."` if it's a development version
  (e.g., `"1.0.1.9000"`)


## `DESCRIPTION` - `License` field


This is from <https://r-pkgs.org/license.html>:

- If you want a permissive license so people can use your code with minimal
  restrictions, choose the MIT license with `use_mit_license()`.
- If you want a copyleft license so that all derivatives and bundles of 
  your code are also open source, choose the GPLv3 license with 
  `use_gpl_license()`.
- If your package primarily contains data, not code, and you want minimal 
  restrictions, choose the CC0 license with `use_cc0_license()`. 
  Or if you want to require attribution when your data is used, choose the 
  CC BY license by calling `use_ccby_license()`.



## `DESCRIPTION` - Who are you?

```
Authors@R:  c(
    person(c("Lucas", "A."), "Nell", email = "lucas@email.com", role = "cre",
           comment = c(ORCID = "LUCAS-ORCID-ID")),
    person(c("Magdalena", "L."), "Warren", email = "maggie@email.com", 
           role = "aut", comment = c(ORCID = "MAGGIE-ORCID-ID")))
```

`role` is typically one or more of the following:

- `"cre"`: creator and maintainer of the package
- `"aut"`: author who made significant contributions
- `"ctb"`: contributor who made relatively small contributions


## `DESCRIPTION` - What does your package need?

### `Imports`

Indicates packages that your package needs to run.

Almost always use `Imports` instead of `Depends` because `Depends` does equivalent
to calling `library()` on all packages listed there.
This clogs up your environment.

```{r use_package, eval = FALSE}
usethis::use_package("<package>")
```


### `Suggests`

Packages that your package can use but are not required.

```{r use_package_Suggests, eval = FALSE}
usethis::use_package("<package>", "Suggests")
```





## Adding a function

Add a new R file to the `R` directory, and write a function inside it.

### **What functions could be useful to you in your work?**

<br> 

Here's my toy example:

```{r hello-fxn, eval = FALSE}
hello <- function(x) {
    paste("Hello, ", x, "!", sep = "")
}
```

## Writing documentation {.smaller .smallcode}

- In RStudio, put your cursor in your function somewhere, then add
  a start to your docs using `Ctrl+Alt+Shift+R`, 
  `Cmd+Option+Shift+R`, or `Code` > `Insert Roxygen Skeleton`.
- After writing your documentation, build your package (`Ctrl+Shift+B` or
  `Cmd+Shift+B`) and try using your function.


```{r hello-fxn-doc, eval = FALSE}
#' Say hello to stuff
#'
#' @param x Single character indicating what you should say hello to.
#'
#' @return A single character saying hello.
#' 
#' @export
#'
#' @examples
#' hello("world")
#' my_hello <- hello("Lucas")
#' 
hello <- function(x) {
    paste("Hello, ", x, "!", sep = "")
}
```

<br>

Any issues? If so, try looking at your `NAMESPACE` file.


## Generating the documentation


> - Our `NAMESPACE` file shows us that our function isn't exported, plus we 
>   don't have aything in our `man` folder that's supposed to contain our docs!
> - We have to generate the documentation for our package! (`Ctrl+Shift+D` or
>   `Cmd+Shift+D`)
> - Document, re-build package, and try your function again.
> - Also try `?<function-name>`



## Adding data

Typically if you create a dataset you want to include with your package...

```{r add-data, eval = FALSE}
cool_data <- data.frame(x = 1:5, y = runif(5))
usethis::use_data(cool_data)
```

Even better is to document how you created the data:

```{r add-data2, eval = FALSE}
usethis::use_data_raw("cool_data")
# Then, inside the newly created `data-raw/cool_data.R`, write:
cool_data <- data.frame(x = 1:5, y = runif(5))
usethis::use_data(cool_data)
```

Also document your datasets! (see <https://r-pkgs.org/data.html>)


## Testing your package

Automated testing makes your package much more robust and helps you avoid
introducing errors as you develop your package.

I highly recommend using the package `testthat` to manage your tests
(see <https://r-pkgs.org/testing-basics.html> and
<https://testthat.r-lib.org/> for more info)



## Example of using a package for a paper

<https://github.com/lucasnell/sauron>


## How could you use this in your projects?


## More info

More on defensive programming and organizing projects in 
the British Ecological Society's *Reproducible Code* found at <https://www.britishecologicalsociety.org/publications/better-science/>

Lots of info on R packages at <https://r-pkgs.org/>

